---
title: "Chicago Crime Patterns (2001–Present)"
subtitle: "Group 8 (STAT 405)"
author:
  - "Preston Rusli"
  - "Naidhruv Iyer"
  - "Christopher White"
date: "23 April 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: true
    embed-resources: true
    smaller: true
    fig-align: center
    css: styles.css
    footer: STAT 405 Final Project
server: shiny
editor: visual
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false
set.seed(405)
library(leaflet)
library(htmltools)
library(scales)
library(stringr)
library(DBI)
library(RSQLite)
library(readr)
library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(DBI)
library(RSQLite)
library(tibble)
library(ggrepel)
library(forcats)

con <- dbConnect(SQLite(), "chicago_crime.sqlite")

#load crime data
crime_raw <- read_csv("chicago_crime.csv", show_col_types = FALSE)
dbWriteTable(con, "crimes", crime_raw, overwrite = TRUE)

#load community area population
pop_raw <- read_csv("acs_community_area_population.csv", show_col_types = FALSE)
dbWriteTable(con, "population", pop_raw, overwrite = TRUE)

# create indexes
invisible(dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_crimes_date ON crimes(Date)"))
invisible(dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_crimes_primary_type ON crimes([Primary Type])"))
invisible(dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_crimes_community_area ON crimes([Community Area])"))
invisible(dbExecute(con, "CREATE INDEX IF NOT EXISTS idx_crimes_arrest ON crimes(Arrest)"))

invisible(dbExecute(con, "DROP TABLE IF EXISTS crimes_clean"))
invisible(dbExecute(con, "
  CREATE TABLE crimes_clean AS
  SELECT *
  FROM crimes
  WHERE Date IS NOT NULL
    AND Year BETWEEN 2001 AND 2025
    AND [Primary Type] IS NOT NULL
    AND [Primary Type] != ''
"))

crime_counts <- dbGetQuery(con, "
  SELECT
    [Primary Type],
    COUNT(*) AS crime_count
  FROM crimes_clean
  GROUP BY [Primary Type]
  ORDER BY crime_count DESC
")

main_crimes <- dbGetQuery(con, "
SELECT
  CASE
    WHEN [Primary Type] IN (
      'THEFT', 'BURGLARY', 'MOTOR VEHICLE THEFT',
      'CRIMINAL DAMAGE', 'CRIMINAL TRESPASS', 'ARSON'
    ) THEN 'Theft / Property'

    WHEN [Primary Type] IN (
      'DECEPTIVE PRACTICE', 'OTHER OFFENSE', 'NON-CRIMINAL'
    ) THEN 'Fraud / Administrative'

    WHEN [Primary Type] IN (
      'BATTERY', 'ASSAULT', 'HOMICIDE',
      'KIDNAPPING', 'INTIMIDATION', 'STALKING'
    ) THEN 'Violent Personal Crime'

    WHEN [Primary Type] = 'ROBBERY'
      THEN 'Robbery'

    WHEN [Primary Type] IN (
      'NARCOTICS', 'OTHER NARCOTIC VIOLATION'
    ) THEN 'Drug Offenses'

    WHEN [Primary Type] IN (
      'WEAPONS VIOLATION', 'CONCEALED CARRY LICENSE VIOLATION',
      'INTERFERENCE WITH PUBLIC OFFICER'
    ) THEN 'Weapons / Law Enforcement'

    WHEN [Primary Type] IN (
      'SEX OFFENSE', 'CRIM SEXUAL ASSAULT', 'CRIMINAL SEXUAL ASSAULT',
      'PROSTITUTION', 'OFFENSE INVOLVING CHILDREN',
      'HUMAN TRAFFICKING', 'PUBLIC INDECENCY', 'OBSCENITY'
    ) THEN 'Sexual / Exploitation Offenses'

    WHEN [Primary Type] IN (
      'PUBLIC PEACE VIOLATION', 'LIQUOR LAW VIOLATION',
      'GAMBLING', 'RITUALISM', 'DOMESTIC VIOLENCE'
    ) THEN 'Public Order / Vice'
  END AS Crime_Group,
  COUNT(*) AS crime_count
FROM crimes_clean
WHERE [Primary Type] IS NOT NULL
GROUP BY Crime_Group
ORDER BY crime_count DESC;")
```

## Title Slide

## Introduction

-   project motivation
-   main questions
-   data source

## Variables and preprocessing

-   key variables
-   cleaning steps

## Weekly Crime Trend

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
#| fig-width: 8
#| fig-height: 6
crime_weekly <- dbGetQuery(con, "
  SELECT
    date(
      date(
        substr(Date, 7, 4) || '-' ||
        substr(Date, 1, 2) || '-' ||
        substr(Date, 4, 2)
      ),
      '-' || ((strftime('%w',
        date(
          substr(Date, 7, 4) || '-' ||
          substr(Date, 1, 2) || '-' ||
          substr(Date, 4, 2)
        )
      ) + 6) % 7) || ' days'
    ) AS week,
    COUNT(*) AS count
  FROM crimes_clean
  WHERE Date IS NOT NULL
    AND CAST(substr(Date, 7, 4) AS INTEGER) BETWEEN 2001 AND 2025
  GROUP BY week
  ORDER BY week
")

crime_weekly <- crime_weekly %>%
  mutate(
    week = as.Date(week),
    count = as.numeric(count)
  )

ggplot(crime_weekly, aes(x = week, y = count)) +
  geom_line(
    aes(color = "Weekly totals"),
    linewidth = 0.35,
    alpha = 0.4
  ) +
  geom_smooth(
    aes(color = "LOESS trend"),
    method = "loess",
    se = FALSE,
    span = 0.25,
    linewidth = 0.8
  ) +
  geom_vline(
    xintercept = as.Date("2020-03-01"),
    linetype = "dashed",
    linewidth = 0.5,
    alpha = 0.7
  ) +
  annotate(
    "text",
    x = as.Date("2020-03-01"),
    y = Inf,
    label = "COVID-19 (Mar 2020)",
    vjust = 1.5,
    hjust = -0.05,
    size = 3
  ) +
  scale_x_date(
    breaks = seq(
      min(crime_weekly$week, na.rm = TRUE),
      max(crime_weekly$week, na.rm = TRUE),
      by = "2 years"
    ),
    labels = date_format("%Y"),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    limits = c(0, NA),
    labels = comma
  ) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Weekly totals" = "gray25",
      "LOESS trend" = "blue"
    )
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, linewidth = 1))) +
  labs(
    title = "Weekly Crimes with Smoothed Trend",
    x = "Year",
    y = "Number of reported crimes (weekly)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top",
    legend.justification = "center",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

-   Reported crime shows a strong long-run decline from the early 2000s into roughly the mid-2010s.
-   There is a visible disruption around 2020–2021 near the COVID-19 marker.
-   The repeating peaks and dips suggest a seasonal cycle on top of the long-run trend.
-   Overall, Chicago crime levels appear shaped by both slow structural change and shorter-term shocks.

## Crime by hour

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
#| fig-width: 8
#| fig-height: 6
hour_counts <- dbGetQuery(con, "
  SELECT
    CASE
      WHEN substr(Date, 21, 2) = 'AM' AND substr(Date, 12, 2) = '12' THEN 0
      WHEN substr(Date, 21, 2) = 'AM' THEN CAST(substr(Date, 12, 2) AS INTEGER)
      WHEN substr(Date, 21, 2) = 'PM' AND substr(Date, 12, 2) = '12' THEN 12
      ELSE CAST(substr(Date, 12, 2) AS INTEGER) + 12
    END AS Hour,
    COUNT(*) AS crime_count
  FROM crimes_clean
  WHERE Date IS NOT NULL
    AND CAST(substr(Date, 7, 4) AS INTEGER) BETWEEN 2001 AND 2025
  GROUP BY Hour
  ORDER BY Hour
")

hour_labels <- c(
  "12AM","1AM","2AM","3AM","4AM","5AM","6AM","7AM","8AM","9AM","10AM","11AM",
  "12PM","1PM","2PM","3PM","4PM","5PM","6PM","7PM","8PM","9PM","10PM","11PM"
)

hour_counts <- hour_counts %>%
  filter(!Hour %in% c(0, 12)) %>%
  mutate(HourLabel = factor(hour_labels[Hour + 1], levels = hour_labels))

ggplot(hour_counts, aes(x = HourLabel, y = crime_count)) +
  geom_col(fill = "lightblue", color = "white") +
  scale_x_discrete(drop = FALSE) +
  labs(
    title = "Crime by Hour of Day",
    x = "Time of Day",
    y = "Number of Crimes"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

-   Reported crime is lowest in the early morning, especially around 4 AM to 6 AM, and rises steadily through the day.
-   Crime activity peaks in the late afternoon and evening, especially around 3 PM and 6 PM to 9 PM, before declining again later at night.

## Monthly heatmap

\[plot\] - 2 takeaways

## Community area crime map
::: columns
::: {.column width="60%"}

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
crime_by_area <- dbGetQuery(con, "
  SELECT
    CAST([Community Area] AS INTEGER) AS Community_Area,
    COUNT(*) AS crime_total_2020_2024,
    COUNT(*) / 5.0 AS crime_avg_per_year
  FROM crimes_clean
  WHERE Date IS NOT NULL
    AND [Community Area] IS NOT NULL
    AND CAST(substr(Date, 7, 4) AS INTEGER) BETWEEN 2020 AND 2024
  GROUP BY CAST([Community Area] AS INTEGER)
  ORDER BY Community_Area
")

areas <- st_read("Boundaries_-_Community_Areas.geojson", quiet = TRUE) %>%
  mutate(
    Community_Area = as.integer(area_numbe),
    comm_name = str_to_upper(str_squish(community))
  )

pop <- read_csv("acs_community_area_population.csv", show_col_types = FALSE) %>%
  transmute(
    comm_name = str_to_upper(str_squish(`Community Area`)),
    population = as.numeric(`Total Population`)
  )

map_df <- areas %>%
  left_join(crime_by_area, by = "Community_Area") %>%
  left_join(pop, by = "comm_name") %>%
  mutate(
    crime_avg_per_year = replace_na(crime_avg_per_year, 0),
    crimes_per_1000_per_year = if_else(
      !is.na(population) & population > 0,
      1000 * crime_avg_per_year / population,
      NA_real_
    )
  ) %>%
  st_make_valid() %>%
  st_transform(4326)

label_text <- paste0(
  "<strong>", str_to_title(map_df$comm_name), "</strong><br>",
  "Crimes per 1,000 residents: ",
  round(map_df$crimes_per_1000_per_year, 1), "<br>",
  "Average annual crimes: ",
  round(map_df$crime_avg_per_year, 1), "<br>",
  "Population: ",
  comma(map_df$population)
)

pal <- colorNumeric(
  palette = rev(viridisLite::plasma(256)),
  domain = map_df$crimes_per_1000_per_year,
  na.color = "#d9d9d9"
)

leaflet(data = map_df) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(crimes_per_1000_per_year),
    weight = 1,
    color = "white",
    opacity = 1,
    fillOpacity = 0.85,
    smoothFactor = 0.2,
    label = lapply(label_text, HTML),
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#333",
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~crimes_per_1000_per_year,
    opacity = 0.85,
    title = "Crimes per 1,000 residents",
    position = "bottomright"
  ) %>%
  fitBounds(-87.95, 41.60, -87.50, 42.05)
```
:::
::: {.column width="40%"}
- Per-capita crime rates give a fairer comparison than raw counts because they account for large differences in community area population.
-   Crime rates vary substantially across Chicago, with a few community areas standing out as concentrated high-rate hotspots rather than crime being evenly distributed.
-   These patterns are descriptive, not causal, since factors like density, transit activity, nightlife, reporting behavior, and policing practices can all affect recorded rates.
:::
:::

## Text Mining Bar Chart

## Arrest model

\[plot\] - key interpretation

## Killer Plot

::: columns
::: {.column width="25%"}
```{r}
sliderInput("year", "Year:", min = 2020, max = 2025, value = 2025, step = 1, sep = "")
```
:::
::: {.column width="75%"}
```{r}
plotOutput("fp_plot", height = "700px")
```
:::
:::

```{r}
#| context: setup
# fingerprint for each year setup ##### 
yearly_finger <- dbGetQuery(con, "
WITH year_counts AS (
  SELECT
    CAST(District AS INTEGER) AS district_num,
    printf('%04d', CAST(Beat AS INTEGER)) AS beat,
    CAST(substr(Date, 7, 4) AS INTEGER) AS year,
    COUNT(*) AS total_crime,
    SUM(CASE
          WHEN [Primary Type] IN (
            'HOMICIDE', 'ROBBERY', 'ASSAULT',
            'BATTERY', 'CRIM SEXUAL ASSAULT', 'KIDNAPPING'
          ) THEN 1 ELSE 0
        END) AS violent_crime,
    SUM(CASE
          WHEN Arrest = 1 OR Arrest = 'TRUE' OR Arrest = 'true' THEN 1 ELSE 0
        END) AS arrests,
    AVG(Longitude) AS lon,
    AVG(Latitude) AS lat
  FROM crimes_clean
  WHERE District IS NOT NULL
    AND Beat IS NOT NULL
    AND Latitude IS NOT NULL
    AND Longitude IS NOT NULL
    AND CAST(District AS INTEGER) IN (1,6,7,9,11,15)
    AND CAST(substr(Date, 7, 4) AS INTEGER) BETWEEN 2020 AND 2025
  GROUP BY CAST(District AS INTEGER), printf('%04d', CAST(Beat AS INTEGER)), CAST(substr(Date, 7, 4) AS INTEGER)
),
volatility AS (
  SELECT
    district_num,
    beat,
    AVG(total_crime) AS mean_yearly_crime,
    CASE
      WHEN AVG(total_crime) = 0 THEN 0
      ELSE sqrt(AVG(total_crime * total_crime) - AVG(total_crime) * AVG(total_crime)) / AVG(total_crime)
    END AS crime_cv
  FROM year_counts
  GROUP BY district_num, beat
)
SELECT
  y.year,
  y.district_num,
  y.beat,
  y.total_crime,
  y.violent_crime,
  y.arrests,
  y.lon,
  y.lat,
  v.mean_yearly_crime,
  v.crime_cv
FROM year_counts y
LEFT JOIN volatility v
  ON y.district_num = v.district_num
 AND y.beat = v.beat
") %>%
  as_tibble() %>%
  mutate(
    violent_rate = violent_crime / total_crime,
    arrest_rate = arrests / total_crime,
    beat_label = paste("Beat", beat),
    district_lab = paste("District", district_num)
  )
# Whorl ridge generator ####
make_whorl_ridge <- function(r, phase = 0, roughness = 0.004, n = 1200) {
  t <- seq(0, 2*pi, length.out = n)

  rad <- r +
    0.06 * sin(2*t + phase) +
    0.03 * sin(5*t + 0.7*phase)

  x <- 0.92 * rad * cos(t)
  y <- 1.18 * rad * sin(t)

  x <- x + 0.22 * exp(-((t - 1.2)^2) / 0.9) - 0.10 * exp(-((t - 4.5)^2) / 0.6)
  y <- y + 0.08 * sin(t - 0.8)

  x <- x + roughness * sin(20*t + phase) + 0.5 * roughness * cos(33*t + 0.4*phase)
  y <- y + roughness * cos(18*t + 0.3*phase)

  tibble(x = x, y = y, t = t)
}

# Smudge / degradation mask ####
apply_smudge <- function(df, strength = 0.22) {
  smudge_base <- df$x + 0.7 * df$y
  smudge_scaled <- scales::rescale(smudge_base, to = c(0, 1), na.rm = TRUE)

  keep_prob <- 1 - strength * pmax(0, smudge_scaled)
  keep_prob[is.na(keep_prob)] <- 1

  keep <- runif(nrow(df)) < keep_prob
  keep[is.na(keep)] <- TRUE

  out <- df
  out[!keep, c("x", "y")] <- NA
  out
}

# Build ridges ####
build_whorl_fingerprint <- function(df_one_district) {
  df_one_district <- df_one_district %>%
    arrange(dist_center) %>%
    mutate(r_val = seq(0.18, 1.48, length.out = n()))

  beat_levels <- levels(df_one_district$beat)
  ridge_list <- vector("list", length(beat_levels))

  for (i in seq_along(beat_levels)) {
    seg <- df_one_district %>% filter(beat == beat_levels[i])

    ridge <- make_whorl_ridge(
      r = seg$r_val[1],
      phase = i * 0.25,
      roughness = seg$roughness[1],
      n = 700
    )

    ridge <- apply_smudge(ridge, strength = 0.18)

    ridge$district_num <- seg$district_num[1]
    ridge$district_lab <- seg$district_lab[1]
    ridge$beat <- as.character(seg$beat[1])
    ridge$total_crime <- seg$total_crime[1]
    ridge$violent_rate_plot <- seg$violent_rate_plot[1]
    ridge$ridge_width <- seg$ridge_width[1]
    ridge$ridge_alpha <- seg$ridge_alpha[1]
    ridge$group_id <- paste0("district", seg$district_num[1], "_ridge", i)
    ridge$x <- ridge$x * 1.12
    ridge$y <- ridge$y * 1.12
    ridge_list[[i]] <- ridge
  }
  bind_rows(ridge_list)
}

# fp_whorl setup ####
precomputed_fp <- vector("list", length = 6)
names(precomputed_fp) <- as.character(2020:2025)
for (yr in 2020:2025) {
  finger_year <- yearly_finger %>% filter(year == yr)
  finger_df2 <- finger_year %>%
    group_by(district_num) %>%
    mutate(
      district_lon = mean(lon, na.rm = TRUE),
      district_lat = mean(lat, na.rm = TRUE),
      dist_center = sqrt((lon - district_lon)^2 + (lat - district_lat)^2)
    ) %>%
    arrange(district_num, dist_center, .by_group = TRUE) %>%
    mutate(
      beat = factor(beat_label, levels = beat_label),
      ridge_width = scales::rescale(rank(violent_rate), to = c(0.5, 4.2)),
      ridge_alpha = scales::rescale(arrest_rate, to = c(0.45, 0.95)),
      crime_cv = dplyr::coalesce(crime_cv, median(crime_cv, na.rm = TRUE)),
      roughness = scales::rescale(crime_cv, to = c(0.001, 0.010)),
      violent_rate_plot = violent_rate
    ) %>%
    ungroup()
  fp_whorl <- finger_df2 %>%
    group_split(district_num) %>%
    lapply(build_whorl_fingerprint) %>%
    bind_rows()
  precomputed_fp[[as.character(yr)]] <- fp_whorl
}
```

```{r}
#| context: server
output$fp_plot <- renderPlot({
  yr <- input$year
  set.seed(405)
  finger_year <- yearly_finger %>% filter(year == input$year)
  req(nrow(finger_year) > 0)
  finger_df2 <- finger_year %>%
    group_by(district_num) %>%
    mutate(
      district_lon = mean(lon, na.rm = TRUE),
      district_lat = mean(lat, na.rm = TRUE),
      dist_center = sqrt((lon - district_lon)^2 + (lat - district_lat)^2)
    ) %>%
    arrange(district_num, dist_center, .by_group = TRUE) %>%
    mutate(
      beat = factor(beat_label, levels = beat_label),
      ridge_width = scales::rescale(rank(violent_rate), to = c(0.5, 4.2)),
      ridge_alpha = scales::rescale(arrest_rate, to = c(0.45, 0.95)),
      crime_cv = ifelse(is.na(crime_cv), median(crime_cv, na.rm = TRUE), crime_cv),
      roughness = scales::rescale(crime_cv, to = c(0.001, 0.010)),
      violent_rate_plot = violent_rate
    ) %>%
    ungroup()
  
  fp_whorl <- precomputed_fp[[as.character(input$year)]]
  core_extra <- finger_df2 %>%
    group_by(district_num, district_lab) %>%
    group_map(~{
      inner_df <- .x %>% slice_tail(n = min(5, nrow(.x)))
  
      bind_rows(lapply(seq_len(nrow(inner_df)), function(j) {
        seg <- inner_df %>% slice(j)
        ridge <- make_whorl_ridge(
          r = c(0.035, 0.055, 0.075, 0.090, 0.105)[j],
          phase = 2 + j * 0.18,
          roughness = seg$roughness * 0.5,
          n = 700
        )
        ridge$district_num <- .y$district_num[1]
        ridge$district_lab <- .y$district_lab[1]
        ridge$beat <- as.character(seg$beat[1])
        ridge$total_crime <- seg$total_crime[1]
        ridge$violent_rate_plot <- seg$violent_rate_plot[1]
        ridge$ridge_width <- max(0.35, seg$ridge_width[1] * 0.8)
        ridge$ridge_alpha <- seg$ridge_alpha[1]
        ridge$group_id <- paste0("district", .y$district_num[1], "_core", j)
        ridge
      }))
    }) %>%
    bind_rows()
  
  fp_whorl <- bind_rows(fp_whorl, core_extra) %>%
    filter(!is.na(district_lab)) %>%
    mutate(
      district_lab = factor(
        district_lab,
        levels = c("District 1", "District 6", "District 7",
                   "District 9", "District 11", "District 15")
      )
    )

  beat_labels <- fp_whorl %>%
    filter(!grepl("_core", group_id)) %>%
    filter(!is.na(x), !is.na(y)) %>%
    group_by(district_lab, group_id, beat, total_crime) %>%
    slice_max(order_by = x, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    group_by(district_lab) %>%
    arrange(desc(total_crime), .by_group = TRUE) %>%
    slice_head(n = 4) %>%
    ungroup()
  if (nrow(beat_labels) == 0) beat_labels <- NULL

  ggplot(
    fp_whorl,
    aes(
      x = x, y = y, group = group_id,
      color = violent_rate_plot,
      linewidth = ridge_width,
      alpha = ridge_alpha
    )
  ) +
    geom_path(lineend = "round", linejoin = "round", na.rm = TRUE) +
    geom_label_repel(
      data = beat_labels,
      aes(x = x, y = y, label = beat),
      inherit.aes = FALSE,
      color = "black",
      size = 2.6,
      fontface = "bold",
      box.padding = 0.16,
      point.padding = 0.06,
      label.size = 0.15,
      min.segment.length = 0,
      segment.color = "#4a3428",
      segment.size = 0.22,
      fill = scales::alpha("white", 0.82),
      seed = 123,
      max.overlaps = Inf,
      show.legend = FALSE
    ) +
    facet_wrap(~ district_lab, ncol = 3, drop = TRUE) +
    coord_equal(xlim = c(-1.55, 1.95), ylim = c(-1.75, 1.70), clip = "off") +
    scale_color_gradient(
      low = "#f6d6d6",
      high = "#8b0000",
      name = "Violence rate",
      labels = scales::percent_format(accuracy = 1)
    ) +
    scale_linewidth(range = c(0.35, 1.8), guide = "none") +
    scale_alpha(range = c(0.45, 0.92), guide = "none") +
    labs(
      title = paste("Crime Fingerprints by Police District,", yr),
      subtitle = "Ridges are ordered by beat distance from each district center"
    ) +
    theme_void()
}, res = 120)
```

## Conclusion

-   main findings
-   limitations
-   future work


```{r}
#| context: setup
onStop(function() {
  if (DBI::dbIsValid(con)) DBI::dbDisconnect(con)
})
```